---
title: "Lab 4"
author: "Seungah Lee"
date: "10/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Data

```{r}
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)

def <- read_excel('cyclonesFootball2019.xlsx', sheet='Defensive')
off <- read_excel('cyclonesFootball2019.xlsx', sheet='Offensive')
bio <- read_excel('cyclonesFootball2019.xlsx', sheet='Biography')

str(def)
str(off)
str(bio)
```


## Part one: Cleaning data

1. The player names and opponent names are factors.

2. All offensive and defensive statistics are numerical. The Weight column in biogrpahy is also numeric. (Optional: Learn to mutate multiple columns systematically using the ?across function in `dplyr)

3. Change the Height column in biogrpahy into numeric. (Hint: Use a unit that works well)

```{r}
defClean <- def %>%
  mutate(Name = as.factor(Name), 
         Opponent_Opponent = as.factor(Opponent_Opponent), 
         across(where(is.character), as.numeric))

offClean <- off %>%
  mutate(Name = as.factor(Name), 
         Opponent_Opponent = as.factor(Opponent_Opponent), 
         across(where(is.character), as.numeric))


bioClean <- bio %>%
  mutate(Name = as.factor(Name)) %>%
  mutate(Weight = as.numeric(Weight))  %>%
  separate(Height, c('Feet', 'Inches'), sep = '-') %>%
  mutate(Feet = as.numeric(Feet) * 12, Inches = as.numeric(Inches)) %>%
  mutate(Height = Feet + Inches) %>%
  select(-c(Feet, Inches))
```


```{r}
str(defClean)
str(offClean)
str(bioClean)
```

<hr>

## Part two: Tidying

1. Reformat the defClean data frame into a tidy format using pivot_longer. The type of statistic (Tackles_Solo, Tackles_ASST, etc) is added as a new key column named stat.

```{r}
defClean0 <- defClean %>%
  pivot_longer(Tackles_Solo:Pass_PB, names_to = 'stat', values_to = 'count')

str(defClean0)
```


2. Compare the distributions of the defensive statistics. What defensive skills are rare?

```{r}
defClean0 %>%
  group_by(stat) %>%
  summarise(defstat = sum(count)) %>%
  arrange(defstat)
```

```{r}
defClean0 %>% ggplot(aes(x = count)) + geom_bar() + facet_wrap(~stat)
```

As shown on the numerical and graphical summary, Turnovers (Turnover_FF, Turnover_FR, Turnover_INT) are seem to be most rare defensive skills.

3. Did ISU have better defense against Iowa or Notre Dame? Answer this question by creating a scatterplot with x- and y-axes being the number of solo tackles (of each player). A large number of solo tackles is an indicator of good defense.

```{r}
defClean0 %>%
  filter(stat == "Tackles_Solo") %>%
  filter((Opponent_Opponent == "Iowa" | Opponent_Opponent == "Notre Dame")) %>%
  pivot_wider(names_from = 'Opponent_Opponent', values_from = 'count') %>%
  select(c(stat, Iowa, `Notre Dame`)) %>%
  ggplot(aes(x = Iowa, y = `Notre Dame`)) + geom_point() + geom_jitter() +
  xlab('Iowa Solo Tackles') +
  ylab('Notre Dame Solo Tackles') +
  ggtitle('ISU Defense Against Iowa and Notre Dame') +
  geom_abline(slope = 1, intercept = 0)
```

The scatterplot shows that ISU had better defense record against Iowa than Notre Dame.

4. Separate the Hometown column into two columns, namely the city and the state. Print the first few rows of these two columns. (Hint: look at the sep= argument of separate)

```{r}
bioClean0 <- bioClean %>%
  separate(Hometown, c('city', 'state'), sep = ',')

head(bioClean0)
```

5. How many players are there from each state?

```{r}
bioClean1 <- bioClean0 %>% 
  group_by(state) %>% 
  summarize(n_players = n()) %>% 
  arrange(desc(n_players))

print(bioClean1)
```

Majority of ISU players are from Iowa, which is 35.

<hr>

## Part three: Joining data frames

1. Find which states do the ISU defensive and offensive players come from. Create an appropriate summary.

```{r}
defFrom <- defClean %>% 
  left_join(bioClean0, by = 'Name') %>% 
  group_by(state) %>%
  select(Name, state) %>% 
  distinct(Name, .keep_all = TRUE) %>%
  summarise(defensiveFrom = n(), .groups = "keep") %>%
  arrange(desc(defensiveFrom))

print(defFrom)
```
```{r}
offFrom <- offClean %>% 
  left_join(bioClean0, by = 'Name') %>% 
  group_by(state) %>%
  select(Name, state) %>% 
  distinct(Name, .keep_all = TRUE) %>%
  summarise(offensiveFrom = n(), .groups = "keep") %>%
  arrange(desc(offensiveFrom))

print(offFrom)
```

2. How does the performance of quarterback Brock Purdy relate to the performance of the ISU team?


```{r}
brock <- offClean %>%
  filter(Name=="Purdy, Brock") %>%
  select(c(Opponent_Opponent, Passing_YDS, Passing_INT, Rushing_YDS))

brock1 <- offClean %>%
  select(Opponent_Opponent, Rushing_YDS)%>%
  group_by(Opponent_Opponent) %>%
  ggplot(aes(y = Opponent_Opponent, weight = Rushing_YDS)) + geom_bar() + 
  ggtitle('Performance of Quarterback Brock Purdy') +
  xlab('Offensive count') +
  ylab('Opponent')

print(brock1)

```

Brock Purdy seems to be a good offensive player especially in the match of ULM.

3. Which ISU player(s) made big progress over the last year? Compare the 2019 and the 2018 data here.

```{r}
def18 <- read_excel('cyclonesFootball2018.xlsx', sheet='Defensive')
off18 <- read_excel('cyclonesFootball2018.xlsx', sheet='Offensive')

defClean18 <- def18 %>%
  mutate(Name = as.factor(Name), 
         Opponent_Opponent = as.factor(Opponent_Opponent), 
         across(where(is.character), as.numeric))

offClean18 <- off18 %>%
  mutate(Name = as.factor(Name), 
         Opponent_Opponent = as.factor(Opponent_Opponent), 
         across(where(is.character), as.numeric))

str(defClean18)
str(offClean18)
```


```{r}
defAll <- inner_join(defClean, defClean18, by ='Name') %>% 
  mutate(Tackles= Tackles_Solo.x - Tackles_Solo.y) %>%
  select(Name, Tackles) %>% 
  arrange(desc(Tackles))

print(defAll) 

ggplot(defAll, aes(x = Name, y = Tackles)) + 
  geom_point() + 
  coord_flip() + 
  xlab('Player') +
  ylab('Defense Record') +
  ggtitle('Progess of Defensive Players') +
  geom_segment(aes(xend = Name, yend = 0))
```

Among the defensive players, Braxton Lewis showed the best progress from 2018 to 2019 based on the Tackles record.

```{r}
offAll <- inner_join(offClean, offClean18, by ='Name') %>% 
  mutate(Receiving = Receiving_YDS.x - Receiving_YDS.y) %>% 
  select(Name, Receiving) %>% 
  arrange(desc(Receiving))

print(offAll)

ggplot(offAll, aes(x = Name, y = Receiving)) + 
  geom_point() + 
  coord_flip() + 
  xlab('Player') +
  ylab('Offense Record') +
  ggtitle('Progess of Offensive Players') +
  geom_segment(aes(xend = Name, yend = 0))
```

Among the offensive players, Deshaunte Jones and Tarique Milton showed the best progress from 2018 to 2019 based on the Receiving record.

